Bias correction in GLMs

### 2021-07-06 bias_correction.rmd
Bias-adjustment for simple simulation

---
title: "Bias correction in GLMs"
author: Bicko, Jonathan & Ben
date: "2021 Jun 21 (Mon)"
compact-title: false
output: "pdf_document"
---

```{R setup, include=FALSE, echo=FALSE}
library(glmmTMB)
library(dplyr)
library(ggplot2)
library(vareffects); varefftheme()
library(ggpubr)
set.seed(1011)

@@

```{R sim_simple}
N <- 1e5
beta0 <- 0.7
betaA <- 0.2
betaW <- 0.5

@@

```{R simple_logistic}
simple_mod <- glm(status ~ age + wealthindex, data = sim_df, family="binomial")
```

@@

```{R simple_coef}
## True beta
true_beta_df <- data.frame(term=c("Intercept", "age", "wealthindex")
	, estimate=c(beta0, betaA, betaW)
)

@@

```{R simple_vareffects} 
# Age
## varpred way
simple_vareff_age <- varpred(simple_mod, "age", isolate=FALSE, modelname="varpred")

# Wealth index
## Not bias adjusted
simple_vareff_wealthindex <- varpred(simple_mod, "wealthindex", isolate=TRUE, modelname="varpred")

@@

```{R simple_logit}
coef_df_logit <- (coef_df
	%>% select(term, estimate, conf.low, conf.high)
	%>% group_by(term)
	%>% summarise_all(plogis)
)
print(coef_df_logit)
```

\subsection{Population averaging}

- Averages of the entire population of the non-focal predictor

```{R pop_ave, fig.cap = "A comparison of population averaged and varpred-based predictions. For `age`, we implement the naive approach to compute the predictions in `popavefun` function and then implement the same in `vareffects` so as to use the centering machineries. In both cases, the population averaging and varpred gives similar estimates. The estimated population average is very close to the observed in the case of `wealthindex` but slightly higher in the case of `age`, see previous paragraph."}

popavefun <- function(mod, focal, non.focal, level=0.95, modelname="Pop. ave", ...) {
	mf <- model.matrix(mod)
	mm <- (mf
		%>% data.frame()
		%>% mutate_at(non.focal, mean)
		%>% as.matrix()
	)
	vc <- vcov(mod)
	linpred <- as.vector(mm %*% coef(mod))
	pse_var <- sqrt(rowSums(mm * t(tcrossprod(data.matrix(vc), mm))))
	z.val <- qnorm(1 - (1 - level)/2)
	pred_df <- (mf
		%>% data.frame()
		%>% select_at(focal)
		%>% mutate(fit = linpred
			, lwr = plogis(fit - z.val*pse_var)
			, upr = plogis(fit + z.val*pse_var)
			, fit = plogis(fit)
			, model = modelname
			, se = NA
		)
	)
	return(pred_df)
}

# simple_vareff_age_pop <- varpred(simple_mod, "age", isolate=TRUE, modelname="Pop. ave")
simple_vareff_age_pop <- popavefun(simple_mod, "age", "wealthindex", modelname = "Pop. ave")

vareff_age <- simple_vareff_age
vareff_age$preds <- do.call("rbind", list(vareff_age$preds, simple_vareff_age_pop))
age_plot <- (plot(vareff_age)
 	+ labs(y="Prob. of improved \n service", colour="Model")
	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
	+ scale_colour_manual(values=c("black", "blue"))
	+ theme(legend.position="bottom")
)

@@

The observed population average is `r true_prop` while the estimated population averages are:

- age: `r mean(simple_vareff_age_pop$fit)`
- wealthindex: `r mean(simple_vareff_wealthindex_adjust$preds$fit)`

@@

```{R model_reff}
reff_mod <- glmmTMB(status ~ age + wealthindex + (1|hhid)
	, data = sim_df
	, family = binomial(link = "logit")
)
```

```{R reff_coef}
## Tidy coef estimates
reff_coef_df <- (broom.mixed::tidy(reff_mod, conf.int=TRUE)
	%>% mutate(term = gsub("\\(|\\)", "", term))
	%>% filter(effect=="fixed")
)
print(reff_coef_df)

@@

```{R reff_vareffects_age} 
## varpred way
reff_vareff_age <- varpred(reff_mod, "age", isolate=FALSE, modelname="varpred")

@@

```{R reff_vareff_wealthindex, fig.cap = "A comparison of population averaged, varpred-based and bias-adjusted predictions. For each of the predictors, the population averaging and varpred gives similar estimates, and slightly off the truth. However, when we apply bias-adjustment, the estimates are very precise (i.e., close to the truth)."}
# Wealth index
## varpred 
reff_vareff_wealthindex <- varpred(reff_mod, "wealthindex", isolate=TRUE, modelname="varpred")

>>>>

---
title: "Bias correction in GLMs"
author: Bicko, Jonathan & Ben
date: "2021 Jun 21 (Mon)"
compact-title: false
output:
  pdf_document:
    extra_dependencies: ["float"]
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(fig.pos = "H", out.extra = "")

library(glmmTMB)
library(dplyr)
library(ggplot2)
library(vareffects); varefftheme()
library(ggpubr)
set.seed(1011)

@@

```{r sim_simple}
N <- 1e5
beta0 <- 0.7
betaA <- 0.2
betaW <- 0.5

@@

```{r simple_logistic}
simple_mod <- glm(status ~ age + wealthindex, data = sim_df, family="binomial")
```

@@

```{r simple_coef}
## True beta
true_beta_df <- data.frame(term=c("Intercept", "age", "wealthindex")
	, estimate=c(beta0, betaA, betaW)
)

@@

```{r simple_vareffects} 
# Age
## varpred way
simple_vareff_age <- varpred(simple_mod, "age", isolate=FALSE, modelname="varpred")

# Wealth index
## varpred - no bias adjustment
simple_vareff_wealthindex <- varpred(simple_mod, "wealthindex", isolate=TRUE, modelname="varpred")

@@

```{r simple_logit}
coef_df_logit <- (coef_df
	%>% select(term, estimate, conf.low, conf.high)
	%>% group_by(term)
	%>% summarise_all(plogis)
)
print(coef_df_logit)
```

\subsection{Variable predictions}

We consider both \textbf{varpred} and population averaging approach; and then introduce bias correction to \textbf{varpred} predictions.

```{r pop_ave_fun}
popavefun <- function(mod, focal, non.focal, level=0.95, modelname="Pop. ave", ...) {
	mf <- model.matrix(mod)
	mm <- (mf
		%>% data.frame()
		%>% mutate_at(non.focal, mean)
		%>% as.matrix()
	)
	vc <- vcov(mod)
	linpred <- as.vector(mm %*% coef(mod))
	pse_var <- sqrt(rowSums(mm * t(tcrossprod(data.matrix(vc), mm))))
	z.val <- qnorm(1 - (1 - level)/2)
	pred_df <- (mf
		%>% data.frame()
		%>% select_at(focal)
		%>% mutate(fit = linpred
			, lwr = plogis(fit - z.val*pse_var)
			, upr = plogis(fit + z.val*pse_var)
			, fit = plogis(fit)
			, model = modelname
			, se = NA
		)
	)
	return(pred_df)
}

```{r simple_vareffects_age} 
## varpred way
simple_vareff_age <- varpred(simple_mod, "age", isolate=FALSE, modelname="varpred")

## Pop. average
simple_vareff_age_pop <- popavefun(simple_mod, "age", "wealthindex", modelname = "Pop. ave")

## Bias adjust
simple_vareff_age_adjust <- varpred(simple_mod, "age", isolate=TRUE, bias.adjust=TRUE, modelname = "Bias adjust")

vareff_age <- simple_vareff_age
vareff_age$preds <- do.call("rbind", list(vareff_age$preds, simple_vareff_age_pop, simple_vareff_age_adjust$preds))
age_plot <- (plot(vareff_age)
 	+ labs(y="Prob. of improved \n service", colour="Model")
	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
	+ scale_colour_manual(values=c("black", "blue", "red"))
	+ theme(legend.position="bottom")
)
```

```{r simple_vareff_wealthindex, fig.cap = "A comparison of population averaged and varpred-based predictions. We also apply bias-adjustemnt to the `vapred` predictions. For `age`, we implement the naive approach to compute the predictions in `popavefun` function and then implement the same in `vareffects` so as to use the centering machineries. In both cases, the population averaging and varpred gives similar estimates. The estimated population average is very close to the observed in the case of `wealthindex` but slightly higher in the case of `age` for the unadjusted estimates (but very low for bias-adjusted). , see previous paragraph."}
# Wealth index
## varpred 
simple_vareff_wealthindex <- varpred(simple_mod, "wealthindex", isolate=TRUE, modelname="varpred")

## Pop. average
simple_vareff_wealthindex_pop <- varpred(simple_mod, "wealthindex", isolate=TRUE, pop.ave=TRUE, modelname="Pop. ave")

## Bias adjust
simple_vareff_wealthindex_adjust <- varpred(simple_mod, "wealthindex", isolate=TRUE, bias.adjust=TRUE, modelname="Bias adjust")

vareff_wealthindex <- simple_vareff_wealthindex
vareff_wealthindex$preds <- do.call("rbind", list(vareff_wealthindex$preds, simple_vareff_wealthindex_pop$preds, simple_vareff_wealthindex_adjust$preds))
wealthindex_plot <- (plot(vareff_wealthindex)
 	+ labs(y="", colour="Model")
	+ geom_hline(yintercept=true_prop, lty=2, colour="grey")
	+ scale_colour_manual(values=c("black", "blue", "red"))
	+ theme(legend.position="bottom")
)

@@

The observed population average is `r true_prop` while the estimated population averages (similar estimates with `varpred`) are:

- age: unadjusted `r mean(simple_vareff_age_pop$fit)`; adjusted `r mean(simple_vareff_age_adjust$preds$fit)`
- wealthindex: unadjusted `r mean(simple_vareff_wealthindex_pop$preds$fit)`; adjusted `r mean(simple_vareff_wealthindex_adjust$preds$fit)`

@@

```{r model_reff}
reff_mod <- glmmTMB(status ~ age + wealthindex + (1|hhid)
	, data = sim_df
	, family = binomial(link = "logit")
)
```

```{r reff_coef}
## Tidy coef estimates
reff_coef_df <- (broom.mixed::tidy(reff_mod, conf.int=TRUE)
	%>% mutate(term = gsub("\\(|\\)", "", term))
	%>% filter(effect=="fixed")
)
print(reff_coef_df)

@@

```{r reff_vareffects_age} 
## varpred way
reff_vareff_age <- varpred(reff_mod, "age", isolate=FALSE, modelname="varpred")

@@

```{r reff_vareff_wealthindex, fig.cap = "A comparison of population averaged, varpred-based and bias-adjusted predictions. For each of the predictors, the population averaging and varpred gives similar estimates, and slightly off the truth. However, when we apply bias-adjustment, the estimates are very precise (i.e., close to the truth)."}
# Wealth index
## varpred 
reff_vareff_wealthindex <- varpred(reff_mod, "wealthindex", isolate=TRUE, modelname="varpred")
